Categories of Optics

### 2018-08-06 paper.tex
A note on states and costates

  Note that the homsets of $\Optic_\otimes$ are given by a coend indexed by a possibly large category. If $\C$ is small then the coends always exist, but if not, their existence is not guaranteed by the cocompleteness of $\Set$. In this case we should be careful to only discuss $\Optic$ categories where we know that the coends exist by some other means, e.g., by exhibiting an isomorphism of $\Optic_\C((S, S'), (A, A'))$ with a set. For all of the examples we give later, we provide such a isomorphism.

\begin{remark}
If $\C$ is a category with finite products, then $\Lens = \Optic_\times$ is exactly the category of lenses described in the introduction, so long as we restrict to optics of shape $(S, S) \hto (A, A)$. We see that optics correspond to pairs of $\fget$ and $\fput$ functions via the following isomorphisms:
\begin{align*}
  \Optic_\times((S, S'), (A, A'))
  &= \int^{M \in \C} \C(S, M \times A) \times \C(M \times A', S') \\
  &\cong \int^{M \in \C} \C(S, M) \times \C(S, A) \times \C(M \times A', S') && \text{(universal property of product)} \\
  &\cong \C(S, A) \times \C(S \times A', S') && \text{(Yoneda reduction)}
\end{align*}
Let $\conc_\times((S, S'), (A, A'))$ denote this set of concrete lenses. Explicitly the isomorphism states that, given an optic $\rep{l}{r} : (S, S') \hto (A, A')$, the corresponding concrete lens is the pair $(\fget, \fput)$, where $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. In the other direction, given $(\fget, \fput)$, the corresponding optic has representative $\rep{[\id_S, \fget]}{\fput}$.

@@

There are some other easy-to-construct optics; specifically, optics out of  and into the monoidal unit $(I, I)$. Such maps in a monoidal category are typically called states and costates.
\begin{proposition}\label{prop:costates}
  The set of costates $(S, S') \hto (I, I)$ is isomorphic to $\C(S, S')$.
\end{proposition}
\begin{proof}
  \begin{align*}
    \Optic_\otimes((S, S'), (I, I))
    &= \int^{M \in \C} \C(S, M \otimes I) \times \C(M \otimes I, S') \\
    &\cong \int^{M \in \C} \C(S, M) \times \C(M, S') \\
    &\cong \C(S, S')
  \end{align*}
  by Yoneda reduction.
\end{proof}

@@

In particular, for any $S \in \C$, the identity $\id_S$ yields an optic $c_S = \rep{\rho_S^{-1}}{\rho_S} : (S, S) \hto (I, I)$, that we will call the \emph{connector}:
%\begin{center}
%  \input{diagrams/connector-full.tikz}
%\end{center}
%Or, again omitting the unitors:
\begin{center}
  \input{diagrams/connector.tikz}
\end{center}

\begin{proposition}
  Suppose the monoidal unit $I$ of $\C$ is terminal. Then the set of states $(I, I) \hto (A, A')$ is isomorphic to $\C(I, A)$.
\end{proposition}
\begin{proof}
  First, note that
  \begin{align*}
    \Optic_\otimes((I,I), (A,A'))
    &= \int^{M \in \C} \C(I, M \otimes A) \times \C(M \otimes A', I) \\
    &\cong \int^{M \in \C} \C(I, M \otimes A).
  \end{align*}
  as $I$ is terminal. The interior of this coend is mute in the contravariant position, so the coend is equal to the colimit of the functor $\C(I, - \otimes A) : \C \to \Set$. But $\C$ has terminal object $I$, so \[\int^{M \in \C} \C(I, M \otimes A) \cong \C(I, I \otimes A) \cong \C(I, A).\]
\end{proof}

@@

  A \emph{teleological natural transformation} $\alpha : F \Rightarrow G$ is a monoidal natural transformation whose components are all dualisable and that is additionally compatible with the dualisation:
  \[
  \begin{tikzcd}
    {(FX)}^* \ar[r, "\cong"]  \ar[d, "(\alpha_X)^*", swap] & F(X^*) \ar[d, "\alpha_{X^*}"] \\
    {(GX)}^* \ar[r, "\cong", swap] & G(X^*)
  \end{tikzcd}
  \]
\end{definition}

@@

\begin{proposition}
  If $f : S \to A$ and $g : A \to S$ are mutual inverses, then $\iota(f, g) : S \hto A$ is a lawful optic, so $\iota$ restricts to a functor $\mathrm{Core}(\C) \to \Lawful_\M$.
\end{proposition}
\begin{proof}
The conditions are easy to check:
\begin{align*}
\outside(\iota(f, g)) &= \outside(\rep{\lambda_A^{-1} f}{g \lambda_{A}}) = g \lambda_{A} \lambda_A^{-1} f = g f = \id_S \\
\twice(\iota(f, g))
&= \repthree{\lambda_A^{-1} f}{\lambda_A^{-1} f g \lambda_{A}}{g \lambda_{A}} \\
&= \repthree{\lambda_A^{-1} f}{\lambda_A^{-1} \lambda_{A}}{g \lambda_{A}} \\
&= \repthree{\lambda_A^{-1} f}{\id_{IA}}{g \lambda_{A}} \\
&= \once(\iota(f, g))
\end{align*}
\end{proof}

\begin{proposition}
  In the case that $\C$ is symmetric monoidal and $\M = \C$ acts by left-tensor, then $\Lawful_\otimes$ is symmetric monoidal with the unswitched tensor.
\end{proposition}
This would of course make no sense with the switched tensor, as the tensor of two objects would no longer be of the form $(X, X)$.
\begin{proof}
Due to the previous proposition, the structure maps of $\Optic_\M$ are all lawful. We just have to check that $\otimes : \Optic_\M \times \Optic_\M \to \Optic_\M$ restricts to a functor on $\Lawful_\M$. Given lawful optics $\rep{l}{r} : S \hto A$ and $\rep{l'}{r'} : T \hto B$ with residuals $N$ and $M$, the first law is easy to check.
\begin{align*}
\outside(\rep{l}{r} \otimes \rep{l'}{r'})
&= \outside(\rep{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(r \otimes r')}) \\
&= (l \otimes l');(M \otimes s_{A,N} \otimes B);(M \otimes s_{A,N} \otimes B');(r \otimes r') \\
&= (l \otimes l');(r \otimes r') \\
&= \id_{S \otimes T}
\end{align*}
For the second, we follow the same strategy as Proposition~\ref{prop:lawful-category}, and show that the action of $\otimes$ on objects respects the coend relations in $\Twoptic_\M(S, A)$ and $\Twoptic_\M(T, B)$.

>>>>

  Note that the homsets of $\Optic_\otimes$ are given by a coend indexed by a possibly large category. If $\C$ is small then the coends always exist, but if not, their existence is not guaranteed by the cocompleteness of $\Set$. In this case we should be careful to only discuss $\Optic$ categories where we know that the coends exist by some other means, e.g., by exhibiting an isomorphism of $\Optic_\C((S, S'), (A, A'))$ with a set. For all of the examples we give later we provide such a isomorphism.

\begin{remark}
If $\C$ is a category with finite products, then $\Lens = \Optic_\times$ is exactly the category of lenses described in the introduction, so long as we restrict to optics of shape $(S, S) \hto (A, A)$. We see that optics correspond to pairs of $\fget$ and $\fput$ functions via the following isomorphisms:
\begin{align*}
  \Optic_\times((S, S'), (A, A'))
  &= \int^{M \in \C} \C(S, M \times A) \times \C(M \times A', S') \\
  &\cong \int^{M \in \C} \C(S, M) \times \C(S, A) \times \C(M \times A', S') && \text{(universal property of product)} \\
  &\cong \C(S, A) \times \C(S \times A', S') && \text{(Yoneda reduction)}
\end{align*}
%Let $\conc_\times((S, S'), (A, A'))$ denote this set of concrete lenses. 
Explicitly the isomorphism states that, given an optic $\rep{l}{r} : (S, S') \hto (A, A')$, the corresponding concrete lens is the pair $(\fget, \fput)$, where $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. In the other direction, given $(\fget, \fput)$, the corresponding optic has representative $\rep{[\id_S, \fget]}{\fput}$.

@@

There are some other easy-to-construct optics; specifically, optics out of  and into the monoidal unit $(I, I)$. Such maps in a monoidal category are typically called states and costates.
\begin{proposition}\label{prop:costates}
  The set of costates $(S, S') \hto (I, I)$ is isomorphic to $\C(S, S')$.
\end{proposition}
\begin{proof}
  \begin{align*}
    \Optic_\otimes((S, S'), (I, I))
    &= \int^{M \in \C} \C(S, M \otimes I) \times \C(M \otimes I, S') \\
    &\cong \int^{M \in \C} \C(S, M) \times \C(M, S') \\
    &\cong \C(S, S')
  \end{align*}
  by Yoneda reduction, so a state $\rep{l}{r} : (S, S') \hto (I, I)$ corresponds to the morphism $rl : S \to S'$, and a morphism $f : S \to S'$ corresponds to the state $\rep{\rho_S^{-1}}{f \rho_S} : (S, S') \hto (I, I)$
\end{proof}

@@

In particular, for any $S \in \C$, the identity $\id_S$ yields an optic $c_S = \rep{\rho_S^{-1}}{\rho_S} : (S, S) \hto (I, I)$ that we call the \emph{connector}:
%\begin{center}
%  \input{diagrams/connector-full.tikz}
%\end{center}
%Or, again omitting the unitors:
\begin{center}
  \input{diagrams/connector.tikz}
\end{center}

\begin{proposition}\label{prop:states}
  Suppose the monoidal unit $I$ of $\C$ is terminal. Then the set of states $(I, I) \hto (A, A')$ is isomorphic to $\C(I, A)$.
\end{proposition}
\begin{proof}
  First, note that
  \begin{align*}
    \Optic_\otimes((I,I), (A,A'))
    &= \int^{M \in \C} \C(I, M \otimes A) \times \C(M \otimes A', I) \\
    &\cong \int^{M \in \C} \C(I, M \otimes A).
  \end{align*}
  as $I$ is terminal. The interior of this coend is mute in the contravariant position, so the coend is equal to the colimit of the functor $\C(I, - \otimes A) : \C \to \Set$. But $\C$ has terminal object $I$, so \[\int^{M \in \C} \C(I, M \otimes A) \cong \C(I, I \otimes A) \cong \C(I, A).\]

  Explicitly, a state $f : I \to A$ in $\C$ corresponds to the optic $\rep{\lambda_A^{-1} f}{!_{I \times A'}} : (I, I) \hto (A, A')$, where $!_{I \times A'} : I \times A' \to I$ is the unique map.
\end{proof}

@@

%  A \emph{teleological natural transformation} $\alpha : F \Rightarrow G$ is a monoidal natural transformation whose components are all dualisable and that is additionally compatible with the dualisation:
%  \[
%  \begin{tikzcd}
%    {(FX)}^* \ar[r, "\cong"]  \ar[d, "(\alpha_X)^*", swap] & F(X^*) \ar[d, "\alpha_{X^*}"] \\
%    {(GX)}^* \ar[r, "\cong", swap] & G(X^*)
%  \end{tikzcd}
%  \]
\end{definition}

@@

We can check when our basic examples of optics are lawful.

\begin{proposition}
If $p = \rep{l}{r} : S \hto A$ is an optic such that $l$ and $r$ are mutual inverses, then $p$ is lawful.
\end{proposition}
\begin{proof}
The conditions are easy to check:
\begin{align*}
\outside(\rep{l}{r}) &= r l = \id_S \\
\twice(\rep{l}{r})
&= \repthree{l}{lr}{r} \\
&= \repthree{l}{\id_{MA}}{r} \\
&= \once(\rep{l}{r})
\end{align*}
\end{proof}

\begin{corollary}\label{cor:iota-lawful}
  If $f : S \to A$ and $g : A \to S$ are mutual inverses, then $\iota(f, g) : S \hto A$ is a lawful optic, so $\iota$ restricts to a functor $\iota : \mathrm{Core}(\C) \to \Lawful_\M$. \qed
\end{corollary}

\begin{proposition}
A costate $p : (S, S) \hto (I, I)$ corresponding to a morphism $f : S \to S$ via Proposition~\ref{prop:costates} is lawful iff $f = \id_S$.
\end{proposition}
\begin{proof}
The first law states that $\outside(p) = \id_S$, so if $p$ is lawful we have
\[  \id_S = \outside(\rep{\rho_S^{-1}}{f \rho_S}) =  f \rho_S\rho_S^{-1} = f \]
On the other hand, if $f  = \id_S$ then $\rep{\rho_S^{-1}}{\rho_S}$ is lawful because its components are mutual inverses.
\end{proof}

\begin{proposition}
  In the case that $\C$ is symmetric monoidal and $\M = \C$ acts by left-tensor, then $\Lawful_\otimes$ is symmetric monoidal with the unswitched tensor.
\end{proposition}
This would of course make no sense with the switched tensor, as the tensor of two objects would no longer be of the form $(X, X)$.
\begin{proof}
Due to Corollary~\ref{cor:iota-lawful}, the structure maps of $\Optic_\M$ are all lawful. We just have to check that $\otimes : \Optic_\M \times \Optic_\M \to \Optic_\M$ restricts to a functor on $\Lawful_\M$. Given lawful optics $\rep{l}{r} : S \hto A$ and $\rep{l'}{r'} : T \hto B$ with residuals $N$ and $M$, the first law is easy to check.
\begin{align*}
\outside(\rep{l}{r} \otimes \rep{l'}{r'})
&= \outside(\rep{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(r \otimes r')}) \\
&= (l \otimes l');(M \otimes s_{A,N} \otimes B);(M \otimes s_{A,N} \otimes B');(r \otimes r') \\
&= (l \otimes l');(r \otimes r') \\
&= \id_{S \otimes T}
\end{align*}
For the second, we follow the same strategy as Proposition~\ref{prop:lawful-category}, and show that the action of $\otimes$ on objects respects the coend relations in $\Twoptic_\M(S, A)$ and $\Twoptic_\M(T, B)$.
